---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rqaem

<!-- badges: start -->
[![R-CMD-check](https://github.com/nccanderson/rqaem/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/nccanderson/rqaem/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

`rqaem` is an R port of the eye-movement
recurrence-quantification-analysis method introduced by Anderson,
Bischof, Laidlaw, Risko and Kingstone (2013). It computes the
recurrence matrix and all the scalar metrics described in the paper
(`rec`, `det`, `revdet`, `lam`, `tt`, `corm`, `meanline`, `maxline`,
`ent`, `relent`, `clusters`), and ports the reference Python
implementation at **bit-for-bit parity** on the bundled test
fixtures.

On top of that it adds three things the originals don't have:

- a **tidy data-frame workflow** — `rqa_by()` consumes long-format
  fixation data and returns a one-row-per-group tibble,
- **sparse storage** for long fixation sequences (`Matrix::dgCMatrix`
  from `n = 64` upward), and
- **bootstrap significance testing** and a **radius-sweep helper**
  for choosing a working radius (Fig. 7 of the 2013 paper).

## Installation

```r
# install.packages("devtools")
devtools::install_github("nccanderson/rqaem")
```

## Quick start

```{r example}
library(rqaem)

# A short scanpath (16 fixations from the originals' TestRqa.py example).
fix <- tibble::tibble(
  x = c(510, 466, 406, 135, 296, 117, 317, 439,
        302, 444, 507, 341, 459, 493, 630, 655),
  y = c(385, 429, 449, 333, 409, 398, 327, 305,
        270, 347, 454, 327, 270, 293, 341, 431)
)

result <- rqa(fix, radius = 64)
result
```

The recurrence matrix is on `result$recmat`. Plot it directly:

```{r recurrence-plot, fig.width = 4, fig.height = 4}
plot_recurrence(result)
```

## Multi-trial workflow

The headline ergonomic win over the originals: feed `rqa_by()` a
long-format data frame and get back one row per trial with all 13
scalar metrics.

```{r rqa_by}
set.seed(1)
eyedat <- data.frame(
  trial = rep(1:4, each = 25),
  x     = runif(100, 0, 1000),
  y     = runif(100, 0, 1000)
)
rqa_by(eyedat, x = x, y = y, by = "trial", radius = 80)
```

## Significance and radius selection

`rqa_bootstrap()` builds a null distribution by resampling. Two null
types: `"shuffle"` (permute fixation order, preserve positions —
breaks temporal structure) and `"uniform"` (draw fresh fixations from
a uniform-on-screen distribution — breaks spatial structure too).

```{r bootstrap}
boot <- rqa_bootstrap(fix, radius = 64, n = 199, seed = 1L)
summary(boot)
```

`radius_sweep()` runs `rqa()` over a grid of candidate radii and
optionally attaches a bootstrap-baseline ribbon. The companion
`autoplot()` method reproduces the radius-selection plot from the
paper:

```{r radius-sweep, fig.width = 6, fig.height = 4}
sweep <- radius_sweep(fix, radii = c(30, 60, 90, 120, 150),
                      bootstrap_n = 99L, seed = 1L)
ggplot2::autoplot(sweep)
```

## Reference

Anderson, N. C., Bischof, W. F., Laidlaw, K. E. W., Risko, E. F., &
Kingstone, A. (2013). Recurrence quantification analysis of eye
movements. *Behavior Research Methods*, **45**(3), 842-856.
<https://doi.org/10.3758/s13428-012-0299-5>

The reference Python and MATLAB implementations were written by
Walter F. Bischof and are bundled under `rqa_original/` in this
repository's workspace.
