---
title: "Getting started with rqaem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rqaem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`rqaem` implements the eye-movement recurrence-quantification-analysis
method of Anderson, Bischof, Laidlaw, Risko and Kingstone (2013) as an
R package. It ports the reference Python implementation at bit-for-bit
parity on the bundled test fixtures, and adds a tidy data-frame
workflow.

This vignette walks through three common entry points:

1. Running [`rqa()`][rqa] on a single fixation sequence.
2. Inspecting the result via [`print()`][print.rqa_result] /
   [`summary()`][summary.rqa_result] /
   [`as.data.frame()`][as.data.frame.rqa_result].
3. Running [`rqa_by()`][rqa_by] over a multi-trial data frame.

```{r setup}
library(rqaem)
```

## One trial

A fixation sequence is just an `n` x 2 matrix or data frame of `(x, y)`
coordinates. The other required argument is `radius`, the threshold
distance (in the same units as the coordinates) within which two
fixations count as recurrent.

```{r}
fix <- rbind(
  c(510, 385), c(466, 429), c(406, 449), c(135, 333),
  c(296, 409), c(117, 398), c(317, 327), c(439, 305),
  c(302, 270), c(444, 347), c(507, 454), c(341, 327),
  c(308, 259), c(459, 270), c(493, 293), c(630, 341)
)

result <- rqa(fix, radius = 64)
result
```

The recurrence matrix itself lives on `result$recmat`; for short
sequences it is a dense base `matrix`, and from `n >= 64` onwards it
switches to a sparse `Matrix::dgCMatrix` so long sequences stay
compact.

`summary()` repeats the metrics and adds a heads-up about any that
weren't computable (for example, `det` is `NA` when no diagonal-line
runs reach the threshold `line_length`):

```{r}
summary(result)
```

`as.data.frame()` flattens the scalar metrics into a one-row tibble.
That is what you want when collecting results across many trials:

```{r}
as.data.frame(result)
```

## Duration-weighted variant

Passing a `duration` vector switches to the duration-weighted
formulation: cells in the recurrence matrix become `dur[i] + dur[j]`
rather than `1`, and line measures sum durations rather than counting
cells.

```{r}
durations <- c(343, 232, 276, 312, 412, 188, 220, 305, 270, 261,
               294, 327, 259, 220, 196, 342)
rqa(fix, radius = 64, duration = durations)
```

## Many trials in a data frame

`rqa_by()` is the headline workflow. Long-format fixation data with a
trial column (and optionally participant, image, etc.) goes in; one
row per group with all 13 scalar metrics comes out.

```{r}
set.seed(1)
eyedat <- data.frame(
  trial = rep(1:4, each = 25),
  x     = runif(100, 0, 1000),
  y     = runif(100, 0, 1000)
)
rqa_by(eyedat, x = x, y = y, by = "trial", radius = 80)
```

Multiple grouping columns work the same way:

```{r}
eyedat$participant <- rep(c("A", "B"), each = 50)
rqa_by(eyedat, x = x, y = y,
       by = c("trial", "participant"), radius = 80)
```

Add a duration column to switch to the weighted variant trial-by-trial:

```{r}
eyedat$duration <- sample(150:500, 100, replace = TRUE)
rqa_by(eyedat, x = x, y = y, duration = duration,
       by = "trial", radius = 80)
```

## What's next

The recurrence-plot helper `plot_recurrence()` and the
bootstrap-significance / radius-selection helpers
(`rqa_bootstrap()`, `radius_sweep()`) ship in later milestones.

## Reference

Anderson, N. C., Bischof, W. F., Laidlaw, K. E. W., Risko, E. F., &
Kingstone, A. (2013). Recurrence quantification analysis of eye
movements. *Behavior Research Methods*, **45**(3), 842-856.
<https://doi.org/10.3758/s13428-012-0299-5>
