---
title: "Getting started with rqaem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rqaem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`rqaem` implements the eye-movement recurrence-quantification-analysis
method of Anderson, Bischof, Laidlaw, Risko and Kingstone (2013) as an
R package. It ports the reference Python implementation at bit-for-bit
parity on the bundled test fixtures, and adds a tidy data-frame
workflow.

This vignette walks through three common entry points:

1. Running [`rqa()`][rqa] on a single fixation sequence.
2. Inspecting the result via [`print()`][print.rqa_result] /
   [`summary()`][summary.rqa_result] /
   [`as.data.frame()`][as.data.frame.rqa_result].
3. Running [`rqa_by()`][rqa_by] over a multi-trial data frame.

```{r setup}
library(rqaem)
```

## One trial

A fixation sequence is just an `n` x 2 numeric matrix or data frame
of `(x, y)` coordinates — one row per fixation. The other required
argument is `radius`, the threshold distance (in the same units as
the coordinates) within which two fixations count as recurrent.

The example below is a 16-fixation scanpath taken from the
originals' `TestRqa.py` example; the coordinates are screen pixels
on a 1000 x 750 display.

```{r}
fix <- tibble::tibble(
  x = c(510, 466, 406, 135, 296, 117, 317, 439,
        302, 444, 507, 341, 308, 459, 493, 630),
  y = c(385, 429, 449, 333, 409, 398, 327, 305,
        270, 347, 454, 327, 259, 270, 293, 341)
)

result <- rqa(fix, radius = 64)
result
```

The recurrence matrix itself lives on `result$recmat`; for short
sequences it is a dense base `matrix`, and from `n >= 64` onwards it
switches to a sparse `Matrix::dgCMatrix` so long sequences stay
compact.

`summary()` repeats the metrics and adds a heads-up about any that
weren't computable (for example, `det` is `NA` when no diagonal-line
runs reach the threshold `line_length`):

```{r}
summary(result)
```

`as.data.frame()` flattens the scalar metrics into a one-row tibble.
That is what you want when collecting results across many trials:

```{r}
as.data.frame(result)
```

## Duration-weighted variant

Passing a `duration` vector switches to the duration-weighted
formulation: cells in the recurrence matrix become `dur[i] + dur[j]`
rather than `1`, and line measures sum durations rather than counting
cells.

```{r}
durations <- c(343, 232, 276, 312, 412, 188, 220, 305,
               270, 261, 294, 327, 259, 220, 196, 342)
rqa(fix, radius = 64, duration = durations)
```

## Many trials in a data frame

`rqa_by()` is the headline workflow. Long-format fixation data with a
trial column (and optionally participant, image, etc.) goes in; one
row per group with all 13 scalar metrics comes out.

```{r}
set.seed(1)
eyedat <- data.frame(
  trial = rep(1:4, each = 25),
  x = runif(100, 0, 1000),
  y = runif(100, 0, 1000)
)
rqa_by(eyedat, x = x, y = y, by = "trial", radius = 80)
```

Multiple grouping columns work the same way:

```{r}
eyedat$participant <- rep(c("A", "B"), each = 50)
rqa_by(eyedat, x = x, y = y, by = c("trial", "participant"), radius = 80)
```

Add a duration column to switch to the weighted variant trial-by-trial:

```{r}
eyedat$duration <- sample(150:500, 100, replace = TRUE)
rqa_by(eyedat, x = x, y = y, duration = duration, by = "trial", radius = 80)
```

## Plots

Three ggplot2-based helpers cover the standard RQA visualisations.

`plot_recurrence()` draws the recurrence matrix as a tile plot. For
binary RQA the cells are filled with a single colour; for the
duration-weighted variant they use a viridis colour scale. The same
plot is also the default for `ggplot2::autoplot()` on an
`rqa_result`.

```{r recurrence-plot, fig.width = 4, fig.height = 4}
plot_recurrence(result)
```

`plot_fixations()` shows the scanpath and optionally draws circles
of radius `radius` around each fixation:

```{r fixations-plot, fig.width = 5, fig.height = 4}
plot_fixations(fix, radius = 64)
```

`plot_fixations_grid()` overlays a regular grid — handy when you are
about to coarsen xy fixations into a categorical sequence:

```{r fixations-grid-plot, fig.width = 5, fig.height = 4}
plot_fixations_grid(fix, grid_size = 80)
```

## Significance testing and radius selection

`rqa_bootstrap()` produces a null distribution for each metric by
resampling. Two null types are available: `"shuffle"` permutes the
fixation order (preserves the spatial distribution; breaks temporal
structure) and `"uniform"` draws fresh fixations from a uniform
distribution on a given screen (breaks both).

```{r}
boot <- rqa_bootstrap(fix, radius = 64, n = 200, seed = 1L)
summary(boot)
```

`radius_sweep()` runs `rqa()` across a grid of candidate radii. When
you pass `bootstrap_n > 0` it also attaches a percentile-bootstrap
ribbon at each radius. The companion `autoplot()` method reproduces
the radius-selection plot from Fig. 7 of Anderson et al. (2013):

```{r radius-sweep-plot, fig.width = 6, fig.height = 4}
sweep <- radius_sweep(
  fix,
  radii = seq(20, 600, by = 20),
  bootstrap_n = 999L,
  bootstrap_type = "uniform",
  screen = c(max(fix$x) - min(fix$x),
             max(fix$y) - min(fix$y)),
  seed = 1L
)
ggplot2::autoplot(sweep)
```

## Reference

Anderson, N. C., Bischof, W. F., Laidlaw, K. E. W., Risko, E. F., &
Kingstone, A. (2013). Recurrence quantification analysis of eye
movements. *Behavior Research Methods*, **45**(3), 842-856.
<https://doi.org/10.3758/s13428-012-0299-5>
